\title{Deep Learning Final Report}

\author{Louis Etienne Kusno\\
Tsinghua University\\
{\tt\small louis.kusno@insa-lyon.fr}
\and
David Qi\\
Tsinghua University\\
{\tt\small qidavid77@gmail.com}
\and
Spencer Xi Zhou\\
Tsinghua University\\
{\tt\small spencer.zhou42@gmail.com}
}

\maketitle

\begin{abstract}
As personalized nutrition becomes increasingly important, especially for individuals managing conditions like diabetes, there is a growing need for recipe generation systems that account for health-related constraints such as glycemic index. Existing models often prioritize fluency and creativity over nutritional suitability. In this work, we fine-tune a pre-trained GPT-2 model on a curated subset of the RecipeNLG dataset containing low-glycemic-index recipes to explore controlled generation of healthier meals. Our dataset is carefully filtered for nutritional quality and manageable token length, allowing the model to learn structured, coherent, and health-compliant outputs. Using conditional generation based on user-provided ingredients, our system produces recipes that are both nutritionally appropriate and syntactically fluent. While qualitative evaluation suggests promising results, challenges remain in ingredient pairing and generalization to rare items. We discuss possible solutions, including semantic constraints and nutritional validation mechanisms. This work highlights the potential of language models in health-aware recipe generation and sets the foundation for more robust, user-driven, and nutritionally validated applications.
\end{abstract}
